%\section{(Graphs) Strongly Connected Components}

In this question, you will write 2 functions on directed graphs.
We assume that key comparisons take $O(1)$ work  and  that graphs are represented as:
\begin{lstlisting}[language=Caml, numbers=none]
type graph = vertexSet vertexTable
\end{lstlisting}


\begin{problem}[10.]
Given a directed graph $G = (V,E)$, its transpose $G^T$ is
another directed graph on the same vertices, with every edge flipped.
More formally, $G^T = (V,E')$, where
$$E' = \{(b,a)~|~(a,b) \in E \}.$$

\ask
Here is a skeleton of an SML definition for \cd{transpose} that computes
the transpose of a graph. Fill in the blanks to complete the implementation.
Your implementation must have $O(|E| \log |V|)$ work and $O(\log^2 |V|)$
span.

\solfin

\begin{lstlisting}[language=Caml,numbers=none]
function transpose (G : graph) : graph = 
let
  val S = vertexTable.toSeq(G)    

  function flip(u,nbrs) = Seq.map ____     (fn v.(v,u))     ____ (vertexSet.toSeq nbrs)

  val ET = Seq.flatten(Seq.map flip S
  val T = vertexTable.____  collect              ____  ET
in
  vertexTable.map ____ vertexSet.fromSeq T               ____
end
\end{lstlisting} 
\end{problem}


\begin{problem}[10.]
A \emph{strongly connected component} of a directed graph $G = (V,E)$ is a subset $S$ of $V$ such
that every vertex $u \in S$ can reach every other vertex $v \in S$ (i.e.,
there is a directed path from $u$ to $v$), and such that no other vertex in
$V$ can be added to $S$ without violating this condition. Every vertex
belongs to exactly one strongly connected component{} in a graph.

\ask
Implement the function:
\begin{lstlisting}[language=Caml,numbers=none]
val scc~:~graph * vertex -> vertexSet
\end{lstlisting}
such that \cd{scc(G,v)} returns the strongly connected component{} containing $v$. You may assume
the existence of a function:
\begin{lstlisting}[language=Caml,numbers=none]
val reachable: graph * vertex -> vertexSet
\end{lstlisting}
such that \cd{reachable(G,v)} returns all the vertices reachable from $v$
in $G$. Not including the cost of \cd{reachable}, your algorithm must
have $O(|E|\log|V|)$ work and $O(\log^2 |V|)$ span.  You might find
\cd{transpose} useful and can assume the given time bounds.


\solfin
\begin{lstlisting}[language=Caml,numbers=none]
fun scc ($G$ : graph, v : vertex) : vertexSet = 
    ____  vertexSet.intersection(reachable(G,v),}                 ____
    ____ reachable(transpose(G,v)))}                              ____
\end{lstlisting}

\end{problem}

